---
title: "Using the CalFluxTools Example Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the CalFluxTools Example Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

CalFluxTools provides a workflow for parsing, quality-controlling,
visualizing, transforming, and analyzing calcium flux plate-reader data. It parses calcium flux data in the
`statAll` format together with plate maps and a parameter manifest, organizes
the data into plate and plate-set objects, applies processing rules such as
masking, parameter filtering, categorical encoding, blank or missing-value
replacement, and baseline-normalized transformations, and supports downstream
outputs including data completeness reports, plate heatmaps, bar charts,
z-prime factor summaries, PCA, and random-forest classification results.

This vignette walks through the analysis of example calcium flux files distributed with
CalFluxTools. The package includes plate-reader output files in `statAll`
format, matching plate maps, and Excel parameter manifests under
`inst/extdata`.

```{r libraries}
library(CalFluxTools)
library(ggplot2)
```

## Locate the Example Files

Use `system.file()` to locate packaged data after CalFluxTools has been
installed. The helper below also works when knitting this vignette from the
source tree.

```{r locate-files}
example_file <- function(...) {
  installed_path <- system.file("extdata", ..., package = "CalFluxTools")
  if (nzchar(installed_path)) {
    return(installed_path)
  }

  source_paths <- c(
    file.path("..", "inst", "extdata", ...),
    file.path("inst", "extdata", ...)
  )
  existing_paths <- source_paths[file.exists(source_paths)]
  if (length(existing_paths) > 0) {
    return(normalizePath(existing_paths[[1]], mustWork = TRUE))
  }

  stop("Could not locate CalFluxTools example data.")
}

simulated_dir <- example_file("Simulated_data")
stemonix_dir <- example_file("Stemonix_05_2022")

list.files(simulated_dir)
list.files(stemonix_dir)
```

The simulated dataset contains one baseline read, one post-dose read, one
plate map, and one manifest describing the intended analysis parameters. This enables us to use the "paired analysis" pipeline provided by the package.

```{r simulated-paths}
baseline_file <- example_file(
  "Simulated_data",
  "Control_plate1_Test1_BL_n001_Pro2_New_T1.statAll"
)
postdose_file <- example_file(
  "Simulated_data",
  "Control_plate1_Test1_30min_n001_Pro2_New_T1.statAll"
)
plate_map_file <- example_file(
  "Simulated_data",
  "Control_Plate_map_1_PC3_New_T2.csv"
)
parameter_file <- example_file(
  "Simulated_data",
  "CalFlux_Parameter_selection_Control_PC3_New_T5.xlsx"
)
```

## Inspect the Parameter Manifest

The manifest stores analysis settings and per-parameter processing rules.
The copy currently included with this source tree records the original
development root directory, so the low-level examples below read the files
directly from `inst/extdata`.

```{r parameter-manifest}
analysis_settings <- openxlsx::read.xlsx(
  parameter_file,
  sheet = "Analysis_Settings",
  colNames = FALSE
)
analysis_settings <- analysis_settings[!is.na(analysis_settings$X1), ]

knitr::kable(
  analysis_settings[, c("X1", "X2", "X3", "X4", "X5")],
  col.names = c("Setting", "Value", "Value 2", "Value 3", "Value 4")
)

parameter_info <- openxlsx::read.xlsx(parameter_file, sheet = "Parameter_Info")
knitr::kable(head(parameter_info[, c("Statistic", "Suggested_Abbreviation", "Use")]))
```

If you want to run the full wrapper from this example data, first create a
new manifest that points at the installed `extdata` directory and the
filenames distributed with the package.

```{r patch-manifest, eval = FALSE}
patched_parameter_file <- tempfile(fileext = ".xlsx")
file.copy(parameter_file, patched_parameter_file, overwrite = TRUE)

manifest <- openxlsx::loadWorkbook(patched_parameter_file)
settings <- openxlsx::read.xlsx(
  patched_parameter_file,
  sheet = "Analysis_Settings",
  colNames = FALSE
)

settings[settings$X1 == "Root_Directory", "X2"] <- paste0(
  normalizePath(simulated_dir),
  .Platform$file.sep
)
settings[
  settings$X1 == "Control_plate1_Test1_BL_n001_Pro2_New_T1.statAll",
  "X2"
] <- "Control_plate1_Test1_BL_n001_Pro2_New_T1.statAll"
settings[
  settings$X1 == "Control_plate1_Test1_30min_n001_Pro2_New_T1.statAll",
  "X2"
] <- "Control_plate1_Test1_30min_n001_Pro2_New_T1.statAll"

openxlsx::writeData(
  manifest,
  sheet = "Analysis_Settings",
  x = settings,
  colNames = FALSE
)
openxlsx::saveWorkbook(manifest, patched_parameter_file, overwrite = TRUE)

calflux_data <- processCalFluxData(patched_parameter_file)
```

Although it is easiest to run the package using the `processCalFluxData` function, lower-level functions are also available for data parsing, transformation and visualization.

## Parse Plate Reads

Use `read_calflux_data()` to parse each `statAll` file into a `plate` S4
object. The metadata reader used by the package is currently internal, so this
example uses `CalFluxTools:::` to attach the plate map in the same way as the
analysis wrapper.

```{r parse-plates}
baseline_plate <- read_calflux_data(baseline_file)
postdose_plate <- read_calflux_data(postdose_file)

baseline_plate <- CalFluxTools:::read_calflux_metadata(
  plate_map_file,
  baseline_plate
)
postdose_plate <- CalFluxTools:::read_calflux_metadata(
  plate_map_file,
  postdose_plate
)

plate_summary <- data.frame(
  read = c("Baseline", "Post-dose"),
  plate_id = c(baseline_plate@plateId, postdose_plate@plateId),
  wells = c(length(baseline_plate@wellIds), length(postdose_plate@wellIds)),
  parameters = c(length(baseline_plate@parameters), length(postdose_plate@parameters))
)
knitr::kable(plate_summary)
```

The plate map provides the well-level sample annotations used for masking,
controls, test samples, concentrations, and plotting.

```{r summarize-plate-map}
plate_map <- baseline_plate@plateAnnotMap

well_type_summary <- as.data.frame(table(plate_map$WellType))
names(well_type_summary) <- c("WellType", "Wells")
knitr::kable(well_type_summary)

compound_summary <- aggregate(
  Well ~ Compound + WellType,
  data = plate_map,
  FUN = length
)
names(compound_summary)[names(compound_summary) == "Well"] <- "Wells"
compound_summary <- compound_summary[
  order(compound_summary$WellType, compound_summary$Compound),
]
knitr::kable(head(compound_summary, 12))
```

## Check Data Completeness

`missingDataReport()` summarizes missing, blank, empty, masked, and usable
wells per parameter. `calculateParamCV()` provides a quick coefficient of
variation screen on the same plate object.

```{r quality-summaries}
missing_report <- missingDataReport(baseline_plate)
missing_report <- missing_report[
  order(missing_report$Missing_Count, decreasing = TRUE),
]
knitr::kable(head(missing_report, 10))

cv_report <- calculateParamCV(baseline_plate)
cv_report <- cv_report[order(cv_report$CV, decreasing = TRUE, na.last = TRUE), ]
knitr::kable(head(cv_report, 10))
```

## Transform and Visualize data

Create a `plateset`, add the baseline and post-dose plates, then compute the
log2 ratio for each parameter. Because `read_calflux_data()` retains `well_id`
as the first column, set `firstDataCol = 2`.

```{r transform-data}
plate_set <- new("plateset")
plate_set <- addPlate(plate_set, baseline_plate, "Baseline")
plate_set <- addPlate(plate_set, postdose_plate, "PostDose")

plate_set <- dataTransform(
  plate_set,
  platePair = c("Baseline", "PostDose"),
  method = "log2ratio",
  firstDataCol = 2
)

names(plate_set@transformedPlateData)

log2ratio <- plate_set@transformedPlateData[[1]]
preview_parameters <- colnames(log2ratio)[seq_len(min(3, ncol(log2ratio)))]

knitr::kable(
  head(
    cbind(Well = baseline_plate@wellIds, log2ratio[, preview_parameters, drop = FALSE]),
    10
  )
)
```

Plotting the transformed values by row and column helps check whether a signal
is spatially patterned across the plate.

```{r transformed-heatmap}
plot_parameter <- "Rise slope S.D. (RFU/s)"

plot_df <- data.frame(
  Well = baseline_plate@wellIds,
  Row = substr(baseline_plate@wellIds, 1, 1),
  Column = as.integer(sub("^[A-P]", "", baseline_plate@wellIds)),
  Value = as.numeric(log2ratio[[plot_parameter]])
)

ggplot(plot_df, aes(x = Column, y = Row, fill = Value)) +
  geom_tile(color = "white", linewidth = 0.1) +
  coord_fixed() +
  scale_y_discrete(limits = rev(LETTERS[1:16])) +
  scale_x_continuous(breaks = seq(1, 24, by = 2)) +
  scale_fill_gradient2(
    low = "#2C7BB6",
    mid = "white",
    high = "#D7191C",
    midpoint = 0,
    na.value = "grey90"
  ) +
  labs(
    title = paste("Log2 ratio:", plot_parameter),
    x = "Column",
    y = "Row",
    fill = "Log2 ratio"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank())
```

## Reuse the Workflow

The same workflow can be applied to the Stemonix May 2022 files in
`inst/extdata/Stemonix_05_2022` by swapping the file paths:

```{r stemonix-manifest}
stemonix_manifest <- data.frame(file = list.files(stemonix_dir))
knitr::kable(stemonix_manifest)
```

For full report generation, use a patched manifest with `processCalFluxData()`.
For focused exploratory work, parse the `statAll` files, attach the plate map,
and call the lower-level QC and transformation helpers shown above.
